From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-06-05 dictionary.tex
Update book BibSpec

\keywords{function-sheaf dictionary, trace of Frobenius, geometrization, categorification, character sheaves,smooth commutative group schemes, finite fields, local fields, algebraic tori, Neron models, Greenberg functor, quasicharacters, quasicharacter sheaves.}

@@

@@

%%%%%%%%%%%%%%% RINGS AND GROUPS %%%%%%%%%%%%%%%
\newcommand{\FF}{{\mathbb{F}}}
\newcommand{\ZZ}{{\mathbb{Z}}}
\newcommand{\NN}{{\mathbb{N}}}
\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\EE}{\mathbb{\bar Q}_\ell}
\newcommand{\OK}{\mathcal{O}_K}
\newcommand{\OL}{\mathcal{O}_L}
\newcommand{\OO}[1]{\mathcal{O}_{#1}}
\newcommand{\bFq}{\bar{k}}
\newcommand{\Fq}{k}
\newcommand{\Fqm}{k_m}
\newcommand{\EEx}{\EE^\times}
\newcommand{\Weil}[1]{\mathcal{W}_{#1}}
\newcommand{\m}{{\mathfrak{m}}}
%%%%%%%%%%%%%%% ALGEBRAIC GROUPS %%%%%%%%%%%%%%%
\newcommand{\Gm}[1]{\mathbb{G}_{\hskip-1pt\textbf{m},#1}}
\DeclareMathOperator{\GL}{GL}
\newcommand{\comp}{\Pi} % Component group
%%%%%%%%%%%%%%% NAMED OPERATORS %%%%%%%%%%%%%%%
\DeclareMathOperator{\Gal}{Gal}
\newcommand{\Frob}[1]{\operatorname{F}_{#1}}
\DeclareMathOperator{\Aut}{Aut}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\ord}{ord}
\DeclareMathOperator{\coker}{coker}
\DeclareMathOperator{\Gr}{Gr}
\DeclareMathOperator{\Irrep}{Irrep}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\Res}{Res}
\DeclareMathOperator{\Nm}{Nm}
\DeclareMathOperator{\trace}{Tr}
\DeclareMathOperator{\obj}{obj}
\DeclareMathOperator{\mor}{mor}
\DeclareMathOperator{\Lang}{Lang}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\Loc}{Loc}
\DeclareMathOperator{\Tot}{Tot}
\newcommand{\gal}[1]{{\operatorname{Gal}\hskip-1pt\left( {\bar #1}/#1 \right)}}
\newcommand{\Spec}[1]{{\operatorname{Spec}\hskip-1pt( #1 )}}
%%%%%%%%%%%% MISCELLANEOUS OPERATORS %%%%%%%%%%%%
\newcommand{\sheafHom}{{\mathscr{H}\hskip-4pt{\it o}\hskip-2pt{\it m}}}
\newcommand{\abs}[1]{{\vert #1 \vert}}
\newcommand{\ceq}{{\, :=\, }}
\newcommand{\tq}{{\ \vert\ }}
\newcommand{\iso}{{\ \cong\ }}
\newcommand{\trFrob}[1]{t_{#1}}
\newcommand{\TrFrob}[1]{T_{#1}}
\newcommand{\lTrFrob}[1]{\TrFrob{#1}^2}
%% Limits
\newcommand{\invlim}[1]{\lim\limits_{\overleftarrow{#1}}}
\newcommand{\dirlim}[1]{\lim\limits_{\overrightarrow{#1}}}
\newcommand{\limit}[1]{\mathop{\textsc{lim}}\limits_{#1}}
\newcommand{\colimit}[1]{\mathop{\textsc{colim}}\limits_{#1}}
%% Fonts for quasicharacter sheaves
\newcommand{\cs}[1]{{\mathcal{#1}}}
\newcommand{\gcs}[1]{{\mathcal{\bar #1}}}
\newcommand{\dualgcs}[1]{\gcs{#1}^\dagger}
\newcommand{\dualcs}[1]{\cs{#1}^\dagger}
%% Categories
\newcommand{\CS}{{\mathcal{C\hskip-0.8pt S}}}
\newcommand{\lCS}{{\CS_\text{flb}}}
\newcommand{\lCSiso}[1]{\lCS(#1)_{/\text{iso}}}
\newcommand{\bCS}{{\CS_0}}
\newcommand{\CSiso}[1]{\CS(#1)_{/\text{iso}}}
\newcommand{\bCSiso}[1]{\bCS(#1)_{/\text{iso}}}
\newcommand{\catname}[1]{\normalfont{\textsf{#1}}}
\newcommand{\Sch}[1]{{\catname{Sch}_{/#1}}}
\newcommand{\QCS}{{\mathcal{QC\hskip-0.8pt S}}}
\newcommand{\QCSiso}[1]{\QCS(#1)_{/\text{iso}}}
%% Labeled items
\makeatletter
\newcommand{\labitem}[2]{
\def\@itemlabel{\textbf{#1}}
\item
\def\@currentlabel{#1}\label{#2}}
\makeatother
%% Shorthand for bars
\renewcommand{\bf}{\bar{f}}
\newcommand{\bg}{\bar{g}}
\newcommand{\bm}{\bar{m}}
\newcommand{\bG}{\bar{G}}
\newcommand{\bH}{\bar{H}}
\newcommand{\brho}{{\bar\rho}}
%% Spacing control
\newcommand{\tight}[3]{\hspace{-#1pt}{#2}\hspace{-#3pt}}
\newcommand{\GxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\bGxG}{\text{$\bar{G} \tight{1}{\times}{1} \bar{G}$}}
\newcommand{\bfxf}{\text{$\bar{f} \tight{1}{\times}{1} \bar{f}$}}
\newcommand{\GxxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\LxL}{\text{$\gcs{L} \tight{0}{\boxtimes}{0} \gcs{L}$}}
\newcommand{\ExE}{\text{$\cs{E}\tight{0}{\boxtimes}{0}\cs{E}$}}
\newcommand{\bExE}{\text{$\gcs{E}\tight{0}{\boxtimes}{0}\gcs{E}$}}
\newcommand{\AxA}{\text{$A \tight{1}{\times}{1} A$}}
\newcommand{\BxB}{\text{$B \tight{1}{\times}{1} B$}}
\newcommand{\GzxGz}{\text{$G^0 \tight{1}{\times}{1} G^0$}}
%% Hyphenation override
\hyphenation{quasi-character}

@@

\begin{abstract}
We consider the category of character sheaves on smooth commutative group schemes $G$ over finite fields $k$ and expand the scope of the function-sheaf dictionary from connected commutative algebraic groups to this setting.
We show that the group of isomorphism classes of character sheaves is an extension of the group of characters of $G(k)$ by a cohomology group determined by the component group scheme of $G$.
As an application, we find a category of sheaves for which isomorphism classes correspond to quasi\-characters of algebraic tori, abelian varieties and unipotent wound groups over non-Archimedean local fields. 
In this geometrization of quasicharacters, our sheaves are local systems on
Greenberg transforms of locally finite type N\'eron models.
\end{abstract}

@@

\tableofcontents

@@

%This paper extends Deligne's function-sheaf dictionary from connected commutative algebraic groups to smooth commutative group schemes over finite fields and then provides an application to the geometrization and categorification of quasicharacters of algebraic tori, abelian varieties and unipotent wound groups over non-Archimedean local fields.

As Deligne explained in \cite{deligne:SGA4.5}*{Sommes trig.}, if $G$ is a connected commutative algebraic group over a finite field $k$, then the trace of Frobenius provides a bijection between the group $G(\Fq)^*$ of $\ell$-adic characters of $G(\Fq)$ and isomorphism classes of rank-one $\ell$-adic local systems $\mathcal{E}$ on $G$ for which 
\begin{equation}\label{introbox}
m^* \cs{E} \iso \cs{E} \boxtimes \cs{E},
\end{equation}
where $m : G\times G\to G$ is the multiplication map.
%
If one wishes to make a category from this class of local systems, one is led to consider morphisms $\cs{E} \to \cs{E}'$ of sheaves which are compatible with particular choices of \eqref{introbox} for $\cs{E}$ and $\cs{E'}$. 
{\it A priori}, the composition $\cs{E} \to \cs{E}' \to \cs{E}''$ of two such morphisms need not be compatible with the choices of \eqref{introbox} for $\cs{E}$ and $\cs{E}''$.
 However, for connected $G$ the isomorphism \eqref{introbox} is unique, if it exists, and there is no impediment to making the dictionary categorical.

If $G$ is a commutative algebraic group over $\Fq$ which is not connected, then the isomorphism \eqref{introbox} need not be unique. In order to track the choice of isomorphism, consider the category $\bCS(G)$ of pairs $(\cs{E},\mu_\cs{E})$ where $\cs{E}$ is a rank-one local system on $G$ and $\mu_\cs{E} : m^*\cs{E} \to \cs{E}\boxtimes\cs{E}$ is a chosen isomorphism of local systems on $G\times G$. 
In this case, the trace of Frobenius provides an epimorphism from isomorphism classes of objects in $\bCS(G)$ to characters of $G(\Fq)$, but the epimomorphism need not be injective; consequently,
every character of $G(\Fq)$ may be geometrized as a pair $(\cs{E},\mu_\cs{E})$ but perhaps not uniquely.
Indeed, it follows from a special case of the main result of this paper that the kernel of the trace of Frobenius $T_{G} : \bCS(G) \to G(\Fq)^*$ trivial if and only if the \'etale group scheme of connected components of $G$ is finite cyclic.
In general, the defect in the dictionary for characters of commutative algebraic groups over finite fields may be addressed with the following observation: if $(\cs{E},\mu_\cs{E})$ and $(\cs{E}',\mu_\cs{E}')$ determine the same character of $G(\Fq)$ then $\cs{E}\iso \cs{E}'$ as local systems on $G$.

In order to extend the dictionary from commutative algebraic groups over finite fields to smooth commutative group schemes $G$ over $\Fq$,
we replace the local system $\cs{E}$ on $G$ with a Weil local system while retaining the extra structure $\mu_\cs{E}$. 
In this way we are led to the category $\CS(G)$ of {\it character sheaves} on $G$ (\S\ref{ssec:category}):
objects in $\CS(G)$ are triples $(\gcs{L}, \mu,\phi)$, where $\gcs{L}$ is a rank-one local system on ${\bar G} := G\times_{\Spec{\Fq}} \Spec{\bFq}$
and $\phi : \Frob{G}^* \gcs{L}\to \gcs{L}$ and $\mu : {\bar m}^* \gcs{L} \iso \gcs{L} \boxtimes \gcs{L}$ are isomorphisms of sheaves satisfying certain compatibility conditions;
morphisms in $\CS(G)$ are then morphisms of Weil sheaves which are compatible with the extra structure.

We begin a study of the category $\CS(G)$ of character sheaves on $G$ by returning to the case when $G$ is a connected commutative algebraic group over $\Fq$ and thus by revisiting Deligne's function-sheaf dictionary (\S\ref{sec:category}).
We consider character sheaves that arise via base change to $\bFq$ from local systems on $G$ (\S\ref{ssec:descentG}) and
those that appear in a pushforward from a constant sheaf along a discrete isogeny $H \to G$ (\S\ref{ssec:discrete_isogenies}).  
While these constructions make sense even for non-connected $G$, in the connected case we show that every character sheaf can be described in both of these ways (\S\ref{ssec:connected}).  
We use this fact to prove that $\TrFrob{G}$ is an isomorphism for connected commutative algebraic groups $G$
and to determine the automorphism groups of character sheaves on such $G$. 
These results are all well known.

Next, we consider character sheaves on \'etale commutative group schemes $G$ over $\Fq$ (\S\ref{sec:etale}).
\'Etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme;
etale group schemes are algebraic groups only when they are finite.
Our key tool for understanding the trace of Frobenius in the \'etale case 
are a reinterpretation of $\CS(G)$ in terms of stalks (\S\ref{ssec:stalks}) and 
a relationship with the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal(\Fq/\Fq)$ is the Weil group for $\Fq$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\CSiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that 
the group homomorphism 
\[
\TrFrob{G}: \CSiso{G} \to G(\Fq)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is finite cyclic.
%The fact that $\TrFrob{G}$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gcs{L}, \mu,\phi)$ and $(\gcs{L}', \mu',\phi')$ determine the same character of $G(\Fq)$ then $(\gcs{L},\phi)$ and $(\gcs{L}',\phi')$ are isomorphic as Weil sheaves.

%The appearance of $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ and the trace of Frobenius gives an isomorphism $\TrFrob{G}: \CSiso{G} \to \Hom(G(\Fq),\EEx)$; on the other hand, if $G$ an etale commutative group scheme over $\Fq$, then $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ measures the kernel of $\TrFrob{G}: \CSiso{G} \to \Hom(G(\Fq),\EEx)$. 
%This observation makes plain the fact that  . . .
%Indeed, quasicharacter sheaves on \'etale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

Having understood character sheaves on commutative smooth group schemes over $\Fq$ in two extreme cases --
connected commutative algebraic groups and \'etale commutative group schemes -- we turn to the general case  (\S\ref{sec:main}), using the component group sequence
\[
0 \to G^0 \to G \to \pi_0(G) \to 0.
\]
Using pullbacks of character sheaves we obtain a diagram
\[
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \CSiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \CSiso{G} \rar \dar{\TrFrob{G}} & \CSiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0.
  \end{tikzcd}
\]
The rows of this diagram are exact (\S\S\ref{ssec:restriction},\ref{ssec:component}), so we may apply the snake lemma to prove the main theorem of the paper,
\begin{theorem*}[{Thm. \ref{thm:snake}}]
If $G$ is a smooth commutative group scheme over $\Fq$ then
\[
\TrFrob{G} : \CSiso{G} \to G(\Fq)^*
\]
is surjective with kernel $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$.
\end{theorem*}
\noindent
It follows that, unless the etale group scheme $\pi_0(\bG)$ is cyclic, the category $\CS(G)$ contains non-trivial character sheaves lying in the kernel of the trace of Frobenius and therefore lost in translation to characters of $G(\Fq)$ and, moreover, that these {\it invisible character sheaves} are parametrized by $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$.

We also illuminate the nature of the category $\CS(G)$ by showing that every morphism in this category is either an isomorphism or trivial, and by showing
\begin{theorem*}[{Thm. \ref{thm:autornaught}}]
If $G$ is a smooth commutative group scheme over $\Fq$ then
\[
\Aut_{\CS{}}(\cs{L}) \iso  \Hh^1(\pi_0(\bG), \EEx)^{\Weil{}}
\]
for all quasicharacter sheaves $\cs{L}$ on $G$.
\end{theorem*}

>>>>

\keywords{function-sheaf dictionary, trace of Frobenius, geometrization, categorification, quasicharacter sheaves, smooth commutative group schemes, finite fields, local fields, algebraic tori, Neron models, Greenberg functor, quasicharacters}

@@

% We make volume bold rather than vol. #
\BibSpec{book}{%
    +{}  {\PrintPrimary}				{transition}
    +{,} { }							{title}
    +{.} { }							{part}
    +{:} { }							{subtitle}
    +{,} { \PrintEdition}					{edition}
    +{}  { \PrintEditorsB}				{editor}
    +{,} { \PrintTranslatorsC}			{translator}
    +{,} { \PrintContributions}			{contribution}
    +{,} { }							{series}
    +{} { \textbf}						{volume}
    +{,} { }							{publisher}
    +{,} { } 							{organization}
    +{,} { }							{address}
    +{,} { \PrintDateB}					{date}
    +{,} { }							{status}
    +{}  { \parenthesize}				{language}
    +{}  { \PrintTranslation}				{translation}
    +{;} { \PrintReprint}				{reprint}
    +{.} { }							{note}
    +{.} {}							{transition}
    +{}  {\SentenceSpace \PrintReviews}	{review}
}

@@

%%%%%%%%%%%%%%% RINGS AND GROUPS %%%%%%%%%%%%%%%
\newcommand{\FF}{{\mathbb{F}}}
\newcommand{\ZZ}{{\mathbb{Z}}}
\newcommand{\NN}{{\mathbb{N}}}
\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\EE}{\mathbb{\bar Q}_\ell}
\newcommand{\OK}{\mathcal{O}_K}
\newcommand{\OL}{\mathcal{O}_L}
\newcommand{\OO}[1]{\mathcal{O}_{#1}}
\newcommand{\bFq}{\bar{k}}
\newcommand{\Fq}{k}
\newcommand{\Fqm}{k_m}
\newcommand{\EEx}{\EE^\times}
\newcommand{\Weil}[1]{\mathcal{W}_{#1}}
\newcommand{\m}{{\mathfrak{m}}}
%%%%%%%%%%%%%%% ALGEBRAIC GROUPS %%%%%%%%%%%%%%%
\newcommand{\Gm}[1]{\mathbb{G}_{\hskip-1pt\textbf{m},#1}}
\newcommand{\comp}{\Pi} % Component group
%%%%%%%%%%%%%%% NAMED OPERATORS %%%%%%%%%%%%%%%
\DeclareMathOperator{\Gal}{Gal}
\newcommand{\Frob}[1]{\operatorname{F}_{#1}}
\DeclareMathOperator{\Aut}{Aut}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\ord}{ord}
\DeclareMathOperator{\coker}{coker}
\DeclareMathOperator{\Gr}{Gr}
\DeclareMathOperator{\Irrep}{Irrep}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\Res}{Res}
\DeclareMathOperator{\Nm}{Nm}
\DeclareMathOperator{\trace}{Tr}
\DeclareMathOperator{\obj}{obj}
\DeclareMathOperator{\mor}{mor}
\DeclareMathOperator{\Lang}{Lang}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\Loc}{Loc}
\DeclareMathOperator{\Tot}{Tot}
\newcommand{\gal}[1]{{\operatorname{Gal}\hskip-1pt\left( {\bar #1}/#1 \right)}}
\newcommand{\Spec}[1]{{\operatorname{Spec}\hskip-1pt( #1 )}}
%%%%%%%%%%%% MISCELLANEOUS OPERATORS %%%%%%%%%%%%
\newcommand{\sheafHom}{{\mathscr{H}\hskip-4pt{\it o}\hskip-2pt{\it m}}}
\newcommand{\abs}[1]{{\vert #1 \vert}}
\newcommand{\ceq}{{\, :=\, }}
\newcommand{\tq}{{\ \vert\ }}
\newcommand{\iso}{{\ \cong\ }}
\newcommand{\trFrob}[1]{t_{#1}}
\newcommand{\TrFrob}[1]{T_{#1}}
\newcommand{\lTrFrob}[1]{\TrFrob{#1}^2}
%% Limits
\newcommand{\invlim}[1]{\lim\limits_{\overleftarrow{#1}}}
\newcommand{\dirlim}[1]{\lim\limits_{\overrightarrow{#1}}}
\newcommand{\limit}[1]{\mathop{\textsc{lim}}\limits_{#1}}
\newcommand{\colimit}[1]{\mathop{\textsc{colim}}\limits_{#1}}
%% Fonts for quasicharacter sheaves
\newcommand{\qcs}[1]{{\mathcal{#1}}}
\newcommand{\gqcs}[1]{{\mathcal{\bar #1}}}
\newcommand{\dualgcs}[1]{\gqcs{#1}^\dagger}
\newcommand{\dualcs}[1]{\qcs{#1}^\dagger}
%% Categories
\newcommand{\QC}{{\mathcal{Q\hskip-0.8pt C}}}
\newcommand{\lQC}{{\QC_\text{flb}}}
\newcommand{\lQCiso}[1]{\lQC(#1)_{/\text{iso}}}
\newcommand{\bQC}{{\QC_0}}
\newcommand{\QCiso}[1]{\QC(#1)_{/\text{iso}}}
\newcommand{\bQCiso}[1]{\bQC(#1)_{/\text{iso}}}
\newcommand{\catname}[1]{\normalfont{\textsf{#1}}}
\newcommand{\Sch}[1]{{\catname{Sch}_{/#1}}}
%% Labeled items
\makeatletter
\newcommand{\labitem}[2]{
\def\@itemlabel{\textbf{#1}}
\item
\def\@currentlabel{#1}\label{#2}}
\makeatother
%% Shorthand for bars
\renewcommand{\bf}{\bar{f}}
\newcommand{\bg}{\bar{g}}
\newcommand{\bm}{\bar{m}}
\newcommand{\bG}{\bar{G}}
\newcommand{\bH}{\bar{H}}
\newcommand{\brho}{{\bar\rho}}
%% Spacing control
\newcommand{\tight}[3]{\hspace{-#1pt}{#2}\hspace{-#3pt}}
\newcommand{\GxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\bGxG}{\text{$\bar{G} \tight{1}{\times}{1} \bar{G}$}}
\newcommand{\bfxf}{\text{$\bar{f} \tight{1}{\times}{1} \bar{f}$}}
\newcommand{\GxxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\LxL}{\text{$\gqcs{L} \tight{0}{\boxtimes}{0} \gqcs{L}$}}
\newcommand{\ExE}{\text{$\qcs{E}\tight{0}{\boxtimes}{0}\qcs{E}$}}
\newcommand{\bExE}{\text{$\gqcs{E}\tight{0}{\boxtimes}{0}\gqcs{E}$}}
\newcommand{\AxA}{\text{$A \tight{1}{\times}{1} A$}}
\newcommand{\BxB}{\text{$B \tight{1}{\times}{1} B$}}
\newcommand{\GzxGz}{\text{$G^0 \tight{1}{\times}{1} G^0$}}
%% Hyphenation override
\hyphenation{quasi-character}

@@

\begin{abstract}
We introduce the category of quasicharacter sheaves on smooth commutative group schemes $G$ over finite fields $k$ and expand the scope of the function-sheaf dictionary from connected commutative algebraic groups to this setting.
We show that the group of isomorphism classes of quasicharacter sheaves is an extension of the group of characters of $G(k)$ by a cohomology group determined by the component group scheme of $G$.
As an application, we find a category of sheaves for which isomorphism classes correspond to quasi\-characters in the case of algebraic tori, abelian varieties and unipotent wound groups over local fields. 
In this geometrization of quasicharacters, our sheaves are local systems on
Greenberg transforms of locally finite type N\'eron models.
\end{abstract}

@@

@@

Motivated by an application to the geometrization and categorification of quasicharacters of $p$-adic tori,
this paper extends Deligne's function-sheaf dictionary from commutative algebraic groups to smooth commutative group schemes over finite fields.

As Deligne explained in \cite{deligne:SGA4.5}*{Sommes trig.}, if $G$ is a connected commutative algebraic group over a finite field $k$, then the trace of Frobenius provides a bijection between the group $G(\Fq)^*$ of $\ell$-adic characters of $G(\Fq)$ and isomorphism classes of rank-one $\ell$-adic local systems $\mathcal{E}$ on $G$ for which 
\begin{equation}\label{introbox}
m^* \qcs{E} \iso \qcs{E} \boxtimes \qcs{E},
\end{equation}
where $m : G\times G\to G$ is the multiplication map.
%
If one wishes to make a category from this class of local systems, one is led to consider morphisms $\qcs{E} \to \qcs{E}'$ of sheaves which are compatible with particular choices of \eqref{introbox} for $\qcs{E}$ and $\qcs{E'}$. 
{\it A priori}, the composition $\qcs{E} \to \qcs{E}' \to \qcs{E}''$ of two such morphisms need not be compatible with the choices of \eqref{introbox} for $\qcs{E}$ and $\qcs{E}''$.
 However, for connected $G$ the isomorphism \eqref{introbox} is unique if it exists, and there is no impediment to making the dictionary categorical.

If $G$ is a commutative algebraic group over $\Fq$ which is not connected, then the isomorphism \eqref{introbox} need not be unique. In order to track the choice of isomorphism, consider the category of pairs $(\qcs{E},\mu_\qcs{E})$ where $\qcs{E}$ is a rank-one local system on $G$ and $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$ is a chosen isomorphism of local systems on $G\times G$. 
In this case, the trace of Frobenius provides an epimorphism from isomorphism classes of pairs $(\qcs{E},\mu_\qcs{E})$ to characters of $G(\Fq)$, but the epimomorphism need not be injective; consequently,
every character of $G(\Fq)$ may be geometrized as a pair $(\qcs{E},\mu_\qcs{E})$ but perhaps not uniquely in the category of pairs. 
While this might appear to be a defect in the dictionary for characters of commutative algebraic groups over finite fields, it is not serious: if $(\qcs{E},\mu_\qcs{E})$ and $(\qcs{E}',\mu_\qcs{E}')$ determine the same character of $G(\Fq)$ then $\qcs{E}\iso \qcs{E}'$ as local systems on $G$.

In order to extend the dictionary from commutative algebraic groups over finite fields to smooth commutative group schemes $G$ over $\Fq$,
we simply replace the local system $\qcs{E}$ on $G$ with a Weil local system while retaining the extra structure $\mu_\qcs{E}$. 
In this way we are led to the category $\QC(G)$ of {\it quasicharacter sheaves} on $G$ (\S\ref{ssec:category}).
Objects in $\QC(G)$ are triples $(\gqcs{L}, \mu,\phi)$, where $\gqcs{L}$ is a rank-one local system on ${\bar G} := G\times_{\Spec{\Fq}} \Spec{\bFq}$
and $\phi : \Frob{G}^* \gqcs{L}\to \gqcs{L}$ and $\mu : {\bar m}^* \gqcs{L} \iso \gqcs{L} \boxtimes \gqcs{L}$ are isomorphisms of sheaves satisfying certain compatibility conditions.
Morphisms in $\QC(G)$ are then morphisms of Weil sheaves which are compatible with the extra structure.

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\Fq$ (\S\ref{sec:etale});
\'etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
Our key tools for understanding the trace of Frobenius in the \'etale case are a reinterpretation of $\QC(G)$ in terms of stalks (\S\ref{ssec:stalks}) and a
relationship with the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal(\Fq/\Fq)$ is the Weil group for $\Fq$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\QCiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that 
the group homomorphism 
\[
\TrFrob{G}: \QCiso{G} \to G(\Fq)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
%The fact that $\TrFrob{G}$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\Fq)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.

Returning to the case of connected commutative algebraic groups $G$, we consider quasicharacter sheaves that arise via base change to $\bFq$ from local systems on $G$ (\S\ref{ssec:descentG}) and
those that appear in a pushforward from a constant sheaf along a discrete isogeny $H \to G$ (\S\ref{ssec:discrete_isogenies}).  While these constructions make sense even for non-connected $G$, in the connected
case we show that every quasicharacter sheaf can be described in both of these ways (\S\ref{ssec:connected}).  We use this fact to prove that $\TrFrob{G}$ is an isomorphism for connected commutative algebraic groups $G$
and to determine the automorphism groups of quasicharacter sheaves on such $G$.

%The appearance of $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ and the trace of Frobenius gives an isomorphism $\TrFrob{G}: \QCiso{G} \to \Hom(G(\Fq),\EEx)$; on the other hand, if $G$ an etale commutative group scheme over $\Fq$, then $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ measures the kernel of $\TrFrob{G}: \QCiso{G} \to \Hom(G(\Fq),\EEx)$. 
%This observation makes plain the fact that  . . .
%Indeed, quasicharacter sheaves on \'etale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

Having understood quasicharacter sheaves on commutative smooth group schemes over $\Fq$ in two extreme cases --
connected commutative algebraic groups and \'etale commutative group schemes -- we turn to the general case, using the component group sequence
\[
0 \to G^0 \to G \to \pi_0(G) \to 0.
\]
Using pullbacks of quasicharacter sheaves (\S\ref{ssec:Frob}), we obtain a diagram
\[
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0.
  \end{tikzcd}
\]
The rows of this diagram are exact (\S\S\ref{ssec:restriction},\ref{ssec:component}), so we may apply the snake lemma to prove the main theorem of the paper,
\begin{theorem*}[{Thm. \ref{thm:snake}}]
If $G$ is a smooth commutative group scheme over $\Fq$ then
\[
\TrFrob{G} : \QCiso{G} \to G(\Fq)^*
\]
is surjective with kernel $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$.
\end{theorem*}
We also illuminate the nature of the category $\QC(G)$ by showing that every morphism in this category is either an isomorphism or trivial, and by showing
\begin{theorem*}[{Thm. \ref{thm:autornaught}}]
If $G$ is a smooth commutative group scheme over $\Fq$ then
\[
\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx),
\]
for all quasicharacter sheaves $\qcs{L}$ on $G$.
\end{theorem*}

As indicated above, our interest in the function-sheaf dictionary for smooth commutative group schemes
over finite fields comes from an application to $p$-adic representation theory,
specifically to quasicharacters (\S\ref{ssec:quasicharacters}) of $p$-adic tori.
However, we found that our method of passing from $p$-adic tori to group schemes over $\Fq$ applies more generally to
any local field $K$ with finite residue field $\Fq$ and any commutative algebraic group $X_K$ that admits a N\'eron model $X$.
This class of groups includes abelian varieties and unipotent $K$-wound groups in addition to the tori we initially considered.
In this paper we show that quasicharacters of $X(K)$ are geometrized and categorified by quasicharacter sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$.
Although not locally of finite type, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and also a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This structure allows us to adapt our work on quasicharacter sheaves to construct (\S\ref{ssec:qc_on_GN}) a category $\QC(X_K)$
of quasicharacter sheaves for $X_K$, which are certain sheaves on $\Gr_R(X)\times_{\Spec{\Fq}} \Spec{\bFq}$, with extra structure.
The ability to generalize the function-sheaf dictionary to non-connected group schemes plays a crucial role in this application.
